---
title: "Weather and Corn Yield Regressions"
author: "Marian Hsieh"
date: "2/25/2022"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(R.matlab)
library(rnassqs)
```

## Weather Data Analysis (script by Nathan Mueller)

### Load the PRISM daily maximum temperatures

```{r tmax data, message = F, results = F, warning = F}
# daily max temperature
# dimensions: counties x days x years
prism <- readMat("data/prismiowa.mat") 

# look at county #1
t_1981_c1 <- prism$tmaxdaily.iowa[1,,1] #pull data out of 3d array
t_1981_c1[366]

# base R plot & ggplot
#plot(1:366, t_1981_c1, type = "1")
ggplot() +
  geom_line(mapping = aes(x=1:366,y=t_1981_c1)) +
  theme_bw() +
  xlab("day of year") + ylab("daily maximum temperature (C)") +
  ggtitle("Daily Maximum Temperature, Iowa County #1")
```

```{r tidying up, message = F, results = F}
# assign dimensions names to tmax matrix
dimnames(prism$tmaxdaily.iowa) <- list(prism$COUNTYFP, 1:366, prism$years)

# converted 3d matrix into a data frame
tmaxdf <- as.data.frame.table(prism$tmaxdaily.iowa)

# re-label the columns
colnames(tmaxdf) <- c("countyfp","day","year","tmax")
tmaxdf <- tibble(tmaxdf)
```

## Temperature trends

### Summer temperature trends: Winneshiek County

```{r summer temp trends, message = F, results = F}
# convert days/years to numeric (from fct to dbl)
tmaxdf$day <- as.numeric(tmaxdf$day)
tmaxdf$year <- as.numeric(as.character(tmaxdf$year))

# tidy data
winnesummer <- tmaxdf %>%
  filter(countyfp==191 & day >= 152 & day <= 243) %>% #after June 1 AND before August 30
  group_by(year) %>%
  summarize(meantmax = mean(tmax))

# plot + trend line
ggplot(winnesummer, aes(x=year,y=meantmax)) +
  geom_point() +
  theme_bw() +
  labs(x="year", y="Tmax (C)") +
  ggtitle("Summer temperature in Winneshiek County")
  geom_smooth(method = lm)

# regression summary
lm_summertmax <- lm(meantmax ~ year, winnesummer)
summary(lm_summertmax)
```

### Winter temperature trends: Winneshiek County

```{r winter temp trends, message = F, results = F}
# change filtered days to winter
winnewinter <- tmaxdf %>%
  filter(countyfp==191 & (day <= 59 | day >= 335) & !is.na(tmax)) %>% #before Feb 29 OR after Dec 1, exclude NAs
  group_by(year) %>%
  summarize(meantmax = mean(tmax))

# plot + trend line
ggplot(winnewinter, aes(x=year,y=meantmax)) +
  geom_point() +
  theme_bw() +
  labs(x="year", y="Tmax (C)") +
  ggtitle("Winter temperature in Winneshiek County")
  geom_smooth(method = lm)

# regression summary
lm_wintertmax <- lm(meantmax ~ year, winnewinter)
summary(lm_wintertmax)
```

### Multiple regression -- Quadratic time trend

```{r quadratic temp trend, message = F, results = F}
# create sq value for years
winnewinter$yearsq <- winnewinter$year^2

# multiple regression
lm_wintertmaxquad <- lm(meantmax ~ year + yearsq, winnewinter)
summary(lm_wintertmaxquad) # positive year, negative yearsq = time trend appears to have slowed

# add fitted values to df
winnewinter$fitted <- lm_wintertmaxquad$fitted.values

# plot + trend line
ggplot(winnewinter) +
  geom_point(mapping = aes(x=year, y= meantmax)) +
  geom_line(mapping = aes(x=year, y=fitted)) +
  theme_bw() +
  labs(x="year", y="Tmax (C)")

```

### Download NASS corn yield data

```{r yield download, message = F, results = F}
# set our API key with NASS
nassqs_auth(key = "EBACEA54-5ED7-3402-8FC5-C2479C946D6B")

# parameters to query on
params <- list(commodity_desc = "CORN", util_practice_desc = "GRAIN", prodn_practice_desc = "ALL PRODUCTION PRACTICES", year__GE = 1981, state_alpha = "IA")

# download
cornyieldsall <- nassqs_yields(params)

# set county and yield as numeric
cornyieldsall$county_ansi <- as.numeric(cornyieldsall$county_ansi)
cornyieldsall$yield <- as.numeric(cornyieldsall$Value)

# clean and filter this dataset
cornyields <- select(cornyieldsall, county_ansi, county_name, yield, year) %>%
  filter(!is.na(county_ansi) & !is.na(yield))
cornyields <- tibble(cornyields)
```


## Assignment

### Question 1a: Extract Winneshiek County corn yields, fit a linear time trend, make a plot. Is there a significant time trend?

```{r question 1a, message = F}
# filter for Winneshiek County
winnecorn <- cornyields %>%
  filter(county_ansi == 191)

# fit model (yield by year)
lm1 <- lm(yield ~ year, winnecorn)
summary(lm1)

# plot + linear trend line (yield by year)
ggplot(winnecorn, aes(x=year,y=yield)) +
  geom_point() +
  theme_bw() +
  xlab("Year") + ylab("Corn yield (bu/acre)") +
  geom_smooth(method = lm)
```

Answer: Yes, there is a significant linear trend between year and yield. 

### Question 1b: Fit a quadratic time trend (i.e., year + year^2) and make a plot. Is there evidence for slowing yield growth? 

```{r question 1b, message = F}
# create sq values for years
winnecorn$year2 <- winnecorn$year^2

# fit model (including years2)
lm2 <- lm(yield ~ year + year2, winnecorn)
summary(lm2) 

# add fitted values to df
winnecorn$fitted <- lm2$fitted.values

# plot + quadratic trend line (yield by year + years2)
ggplot(winnecorn) +
  geom_point(mapping = aes(x=year, y=yield)) +
  geom_line(mapping = aes(x=year, y=fitted)) +
  theme_bw() +
  xlab("Year") + ylab("Corn yield (bu/acres)")
```

Answer: No, there is no evidence of slowing yield. Adding $year^2$ does not improve the model's fit, and does not show a significant negative coefficient for slowing yield. 

### Question 2 -- Time Series: Let's analyze the relationship between temperature and yields for the Winneshiek County time series. Use data on yield and summer avg Tmax. Is adding year or Tmax^2 to your model helpful? Make a plot and interpret the results.

```{r question 2, message = F}
# merge summer temp (1981-2018) + yield (1981-2021) 
winnety <- merge(winnesummer, winnecorn, by = "year")
  
# fit model (yield by mean max temp)
lm3 <- lm(yield ~ meantmax, winnety)
summary(lm3)

# fit model (including year)
lm4 <- lm(yield ~ meantmax + year, winnety)
summary(lm4)

# create sq values for tmax, fit model (including tmax2), add fitted values to df
winnety$meantmax2 <- winnety$meantmax^2
lm5 <- lm(yield ~ meantmax + meantmax2, winnety)
summary(lm5)
winnety$fitted <- lm5$fitted.values

# plot + quadratic trend line (lm5: yield by meantmax + meantmax2)
ggplot(winnety) +
  geom_point(mapping = aes(x=meantmax, y=yield)) +
  geom_line(mapping = aes(x=meantmax, y=fitted)) +
  theme_bw() +
  xlab("Mean maximum temperature (°C)") + ylab("Corn yield (bu/acre)")
```

Answer: Yes, adding year and $Tmax^2$ to the model are helpful and show a significant quadratic relationship between yield and average maximum summer temperature. There was no significant linear relationship between the two.

### Question 3 -- Cross-Section: Analyze the relationship between temperature and yield across all counties in 2018. Is there a relationship? Interpret the results.

```{r question 3, message = F}
# filter yield for 2018
corn2018 <- subset(cornyields, year == 2018)

# filter summer temp for 2018 (all counties)
tmax2018 <- tmaxdf %>%
  filter(year == 2018 & day >= 152 & day <= 243) %>%
  group_by(countyfp) %>%
  summarize(meantmax = mean(tmax))

# merge yield + summer temp for 2018
ct2018 <- merge(corn2018, tmax2018, by.x = "county_ansi", by.y = "countyfp")

# fit model (yield by mean max temp)
lm6 <- lm(yield ~ meantmax, ct2018)
summary(lm6)

# add sq values to meantmax, fit model (including meantmax2), add fitted values to df
ct2018$meantmax2 <- ct2018$meantmax^2
lm7 <- lm(yield ~ meantmax + meantmax2, ct2018)
summary(lm7)
ct2018$fitted <- lm7$fitted.values

# plot + quadratic trend line (lm7: yield by meantmax + meantmax2)
ggplot(ct2018) +
  geom_point(mapping = aes(x=meantmax, y=yield)) +
  geom_line(mapping = aes(x=meantmax, y=fitted)) +
  theme_bw() +
  xlab("Mean maximum temperature (°C)") + ylab("Corn yield (bu/acre)")
```

Answer: There is not a significant linear relationship between yield and average maximum temperature, as 
$p=0.0631$ and $R^2=0.02687$. There is a significant quadratic relationship between yield and temperature in all counties in 2018, peaking around 28°C ($p=0.00239$ and $R^2=0.1124$).

### Question 4 -- Panel: One way to leverage multiple time series is to group all data into what is called a "panel" regression. Convert the county ID code ("countyfp" or "county_ansi") into factor using as.factor, then include this variable in a regression using all counties' yield and summer temperature data. How does the significance of your temperature coefficients (Tmax, Tmax^2) change? Make a plot comparing actual and fitted yields and interpret the results of your model.

```{r question 4, message = F}
# filter for summer temps (all counties, all years)
tmaxsummers <- tmaxdf %>%
  filter(day >= 152 & day <= 243) %>%
  group_by(countyfp, year) %>%
  rename(county_ansi = countyfp) %>%
  summarize(meantmax = mean(tmax))
        
# merge summer temps + yield (all counties, all years)
alltempcorn <- merge(tmaxsummers, cornyields, by = c("county_ansi", "year"))

# convert county_ansi into factor
alltempcorn$county_ansi <- as.factor(alltempcorn$county_ansi)

# add sq values to meantmax, fit model (including meantmax2)
alltempcorn$meantmax2 <- alltempcorn$meantmax^2
lm8 <- lm(yield ~ meantmax + meantmax2, alltempcorn)
summary(lm8)

# fit model (including county), add fitted values to df
lm9 <- lm(yield ~ meantmax + meantmax2 + county_ansi, alltempcorn)
summary(lm9)
alltempcorn$fitted <- lm9$fitted.values

# plot of actual vs fitted yields
ggplot(alltempcorn, aes(x=yield, y=fitted)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "grey") +
  xlim(10,240) + ylim(10,240) +
  theme_bw() +
  xlab("Actual corn yield (bu/acres)") + ylab("Fitted corn yield (bu/acres)")
```

Answer: Adding county improves the explanatory power of the model from $R^2=0.1943$ to $R^2=0.2382$. However, the significance of Tmax ($1.68^2$) and $Tmax^2$ (-3.13) don't change between models, similarly $<2^-16$ in both. The coefficients also do not change significantly between models either. 

### Question 5 -- Soybeans: Download NASS data on soybean yields and explore either a time series relationship for a given county, the cross-sectional relationship for a given year, or a panel across all counties and years.

```{r question 5 download, message = F, results = F}
# parameters to query on
params <- list(commodity_desc = "SOYBEANS", prodn_practice_desc = "ALL PRODUCTION PRACTICES", year__GE = 1981, state_alpha = "IA")

# download
soyyieldsall <- nassqs_yields(params)
```

```{r question 5 time series analysis (Keokuk), message = F}
# clean and filter, set county and yield to numeric
soyyields <- soyyieldsall %>%
  select(county_ansi, county_name, Value, year) %>%
  mutate(county_ansi = as.numeric(county_ansi), yield = as.numeric(Value)) %>%
  filter(!is.na(county_ansi) & !is.na(yield))
soyyields <- tibble(soyyields)

# filter yield & temp for Keokuk County (ansi = 107), merge datasets (1981-2018)
keosoy <- soyyields %>%
  filter(county_ansi == 107)
keosummer <- tmaxdf %>%
  filter(countyfp == 107 & day >= 152 & day <= 243) %>%
  group_by(year) %>%
  summarize(meantmax = mean(tmax), meantmax2 = meantmax^2, year2 = year^2)
keoty <- merge(keosoy, keosummer, by = "year")
  
# check model (yield by year)
lm10 <- lm(yield ~ year, keoty)
summary(lm10)

# check model (including year2), add fitted values to df; decided to focus on year and not include meantmax
lm11 <- lm(yield ~ year + year2, keoty)
summary(lm11)
keoty$fitted <- lm11$fitted.values

# plot + quadratic trend line (lm11: yield by year + year2)
ggplot(keoty) +
  geom_point(mapping = aes(x=year, y=yield)) +
  geom_line(mapping = aes(x=year, y=fitted)) +
  theme_bw() +
  xlab("Year") + ylab("Soybean yield (bu/acre)") +
  ggtitle("Soybean yield by year in Keokuk County, IA (1981-2018)")
```

Answer: This is a time series analysis of soybean yield in Keokuk County from 1981-2018. There is a significant relationship between year and soybean yield, both when fitted linearly and quadratically. The explanatory power of the quadratic model ($R^2=0.4615$) is just slightly stronger than the linear model ($R^2=0.4597$).

### Bonus #1: Find a package to make a county map of Iowa displaying some sort of information about yields or weather. Interpret your map.

```{r bonus 1, message = F}

```

Answer:

### Bonus #2: Challenge question - map trends in corn yields by county across Iowa. Interpret your map.

```{r bonus 2, message = F}

```

Answer:
